On Devon Allen's Disqualification at the 2022 Track and Field World Championship

### 2023-04-20 Manuscript/manuscript.tex
edits/comments section 1-2

\jy{The bib file needs quality control. The bib keys need to have a consistent
  style. I suggest using the keys from google scholar.}

\jy{Practice my writing tips in the stat-wriging notes: chapter 2 on
  latex. E.g., no vertical/horizontal lines in professional tables. Use the
  tables in my template.}

@@

\jy{use the following structure and flow:
  1) A paragraph to tell the story;
  2) A paragraph on informal statistical analysis on the internet/social media;
  3) A paragraph on scholarly articles on reaction time in the literature;
  4) A paragraph on what's new: methods; results.
  5) A paragraph of roadmap.}

\jy{could condense a bit}
%1
After coming in third in the 110 meter hurdles at the United States Track and 
Field Championships in Eugene Oregon at the end of June, Allen was expected to 
be a contender for the World Championships, which were scheduled for two months 
later, also in Eugene. At the World Championships, Allen made it through the 
heats and semifinals to reach the finals while posting reaction times of 0.123
in the heats and 0.101 in the semifinal.  Allen, who attended the University
of Oregon (Location of US World Championship and the World Championship) was
considered to be a favorite for the 110 meter hurdle after running 12.84 seconds
in early June; only 0.04 seconds away from the world record \citep{wa2022preview}.
Thus his disqualification by reacting in 0.099 seconds instead of 0.01 was met
with shock from both Allen and fans alike.  NBC, who was broadcasting the
championships in the United States, later uploaded a video to their YouTube
channel that show the disqualification and can be found here:
\url{https://www.youtube.com/watch?v=D6NXTMo-1yM}. The crowd erupted with jeers when
they found out that Allen was disqualified, angry at the result.  Had Allen 
reacted just 0.001 seconds slower he would have been allowed to compete and been 
spared waiting another two years until the next World Track and Field Championship.
An investigation into reaction times is also important because reaction times have
been shown to affect the overall time of the race \citep{delalija2008reaction}.
Thus, faster reaction times lead to faster overall times; which is the goal of
every sprinter who steps onto the track.

%2
Outside the blog posts from \url{www.LetsRun.com}, there have not been any published
papers regarding Devon Allen's disqualification in particular.  This is not
particularly surprising given that his disqualification occurred only last August.
Letsrun is a website that is part message board and part news.  The message board,
which functions similarly to Reddit is very active during important running
events, which include during the finals of major events such as
the Olympics and the World Track and Field Championships.  However, it is the
news section of the website that generated articles such as ``Was Devon Allen
Screwed? There's At Least a 99\% That He Was" and ``The Data Keeps Pouring In and
It Continues to Look Bad For World Athletics and Great For Devon Allen".  Robert
Johnson, who created and runs the Letsrun website argued that there was an error
with the timing equipment at the 2022 Track and Field World Championship and
cited graphs, statistics such as median reaction time, and compared the reaction
times at the 2022 Track and Field World Championship \citep{johnson2022devon}.

\jy{break into two paragraphs}
%3
Other studies have looked at minimum reaction times such as \citet{pain2007sprint}  which looked at
reaction times for nine male athletes and analyzed their reaction times.  The
fastest athlete of the nine, had on average (even with their two fastest time 
removed) a reaction time of 0.087 seconds and a standard deviation of 0.004
seconds.  Two other athletes were able to react similarly when under certain
muscular tightening conditions.  In 2009, IAAF commissioned a study to examine
the disqualification barrier. It was concluded that the disqualification 
barrier be decreased to allow sprinters to
react faster, and suggested 0.08 or 0.085 seconds as possible new thresholds.
Additionally, it was recommended to IAAF that high speed cameras be used to
change the reaction time criteria from pushing off on the block to body movement.
\citep{komi2009iaaf}. This could provide a more accurate representation of 
reaction time, and thus the possibility of a false disqualification would be lowered.
The possibility of using a video-based medium has been proved to be a reliable
measure of reaction time \citep{mudric2015evaluation}.

\jy{Not Zhang's. There are coauthors. Consider using citet when appropriate in this paragraph.}

A 2012 follow-up study to Komi's by \citet*{pilianidis2012start} found
significant differences between reaction times at several world championships
between 1997 and 2009 for the men's 110 meter hurdles.  Another important
conclusion was that for reaction times did not decrease over the twelve years
covered by the research in the study \citep{pilianidis2012start}.  This contradicts
what was found in a 2021 study that concluded the year of competition has a
significant effect on reaction times \citep{zhang2021correlation}.  Their
analysis looked at data from the 2011-2019 World Championships across multiple
sprint events and concluded that from 2011 to 2019, reaction times got
faster. However, not  all studies have been in favor of a decrease in the
reaction time barrier. \citet*{brosnan2017effects} in 2016 argue for increasing 
the reaction time barrier to 0.115 seconds, and cited evidence from European 
Championships and World Championships in their article.  A study looking 
at the 2008 Olympics likewise supported the conclusion that athletes could not
react in as little as 0.1 seconds \citep{lipps2011implications}.  Another study
concluded that visual reaction times to the starting gun may be faster by 0.007
seconds than the IAAF method, which is from force on block starts 
\citep{holmes2018method}.  While this may seem trivial, Allen was disqualified
only by 0.01 seconds. It is also worth noting that one study looking at data 
from various IAAF meets found that the fastest reaction times for men were found
between the ages of 26 and 29 \citep{tonnessen2013reaction}.  This is notable 
because Allen was 27 years old at the 2022 Track and Field World Championship 
and in his physical peak.

%How barrier is picked by regulator

%4
\jy{I think you have one objective but two methods to approach it.}
The objective to this paper is to analyze the reaction times at the 2022
World Track and Field Championship to check for any abnormalities, and we approach
this two different ways.  Our first method is to try to model the data, which is
the reaction times of every non-disqualification since the 1999 World Championships. 
The model we use is a linear model, following a gamma distribution with a random
effect, which will be referred to as the venue effect.  The venue effect is
reflective of how significant the changes in the reaction time is for each venue,
i.e. the venue effect is  responsible for the variation between years.  
The second method of this paper is to compare data for athletes who have competed
at additional competitions to see how their reaction times compare from previous
competitions to this one. First we can look at the reaction times of athletes who
competed at the World Track and Field Championship in both 2019 and 2022 and
compare the times, and secondly we can look at how athletes performed compared
to how they did at their national/qualifiers meet. In order to reduce any chances
of athletes improving reaction time, we can also look at data from qualifying meets.
Almost every athlete had to qualify by proving they are elite at the national level, 
usually by coming in at least fourth. Thus, data was compiled on athletes who competed
at the 2022 World Track and Field Championships on what these athletes also ran 
at their country's national championship. If there was any argument that athletes
have improved their reaction time from 2019 to 2022, it should now be considered
nearly null, as the national competitions took place between June and July 2022 
and the World Championships were in August. The equipment used to measure reaction
time cannot be guaranteed to be the same from national to international
competition, but the time span between meets has been reduced considerably.

%Simplify contribution to one paragraph: collect data and formal statistical
%analysis such as model used
%Move a lot of this to methods section

%5
The rest of this paper is as follows: Section~\ref{sec:Data} describes how data was
collected and begins to detail the GLMM model which is further developed in the
Section~\ref{sec:Methods}.  The signed-rank test analysis for the same athletes is
also included in Section~\ref{sec:Methods}.  Section~\ref{sec:Results}
section provides the probabilities from the methods section and the paper ends 
by questioning the reaction time barrier.  Lastly, section~\ref{sec:Discussion}
discusses the impact and limitations of the paper.

@@

\jy{Describe you data types; give details on sources; introduce notations for
  both types of data
  present exploratory plots.}
There are two types of data that is explored throughout this paper: data from
1999-2022 for every world championship, and data from athletes who competed
at both the 2019 and 2022 World Championships and 2022 national competitions
and the 2022 World Championship.  The first data set is used for the GLMM portion
of the paper, while the second data set is used for the rank based comparisons.

\subsection{World Championship}
Data was taken from the World Athletics and covers the men's 110 meter hurdles 
and the women's 100 meter hurdles from 1999 to 2022.  The variable of prediction 
interest is reaction time, and the predicting variables are year, stage of 
competition (heats, semifinal, final), total time, and gender.  
The data can be found in the appendix~\ref{sec:Appendix}. Factors such as name, 
country, and identification number were not considered important for the 
purposes of this research besides the times for Devon Allen and the other 
United States athletes mentioned above.  The variable ``Stage"
refers to whether the observation occurred during the ``heats" (Preliminary round) which
is denoted by ``H" in the data, ``S" for ``semifinals" (Of which there are either 2 or 3 
heats each year), and ``F" for finals. The variable ``Gender" has levels ``M" and ``F" 
which stand for ``male" and ``female" respectively.  The variable ``Batch" takes a 
numerical value and every heat of every round from 1999 to 2022 was assigned it's
own number.  This was added because of suspected issues that reaction time may be
correlated with other runners reacting quickly (one runner reacting quickly may
result in other runners reacting quickly).  There is no statistical significance to
how the batches were labeled, they were ordered chronologically by meet but
reverse chronologically by year (Batch 1 is a preliminary Heat of 2022 and Batch
111 corresponds to the 1999 finals). Included are
box plots of the reaction times of the men's finals and the men's semifinals and
finals pooled together.  We will refer to the latter as the ``pooled data"
throughout the paper.  The heats or preliminary rounds were not studied because
athletes react faster as the competition progresses.  Athletes react faster in
the finals than the semifinals which are faster than the heats 
\citep{zhang2021correlation}.  If we want to determine if Allen's reaction time
is reasonable, then we thus need to look at the men's final data, with the men's
pooled data being included in order to increase the sample size.  In 2022, there
were only five data points due to two disqualifications and one athlete who 
qualified but did not compete.
The data that looked at in the paper is best visualized using a box plot as show 
in Figure~\ref{fig:Boxplots}.  It is clear that the 2022 box plot for both the
men's finals and semifinals data is lower (indicating a faster reaction time)
than in previous years.  The median for the men's pooled data in 2022 was 0.129
and the median for the men's final data in 2022 was 0.126.  For comparison,
Brosnan found reaction times of 0.156 seconds for men when looking at data from
1999 to 2014 \citep{brosnan2017effects}.

Something that is extremely important to note in the Data is the inconsistency 
of World Athletics in terms of how they penalize false starts.  From 2007-2009,
the IAAF (the former name for World Athletics, and the governing body for the 
World Track and Field Champions) instituted a rule change that allowed one 
warning false start before a sprinter was disqualified \citep{iaaf2009falsestart}  This 
essentially gave sprinters a second chance and reduced the penalty for a false 
start.  There were 18 male and 7 female false starts at the 2007 World 
Championships, 18 male and 7 female starts at the 2009 World Championships. 
Starting in 2011, the rule was scrapped, and the old policy which automatically 
disqualifies runners who false start was reinstituted.  The effect? At the 2011 
World Championship there were 6 male disqualifications and 4 female 
disqualifications \citep{iaaf2009falsestart}. By returning to the harsher policy and 
cracking down on false starts, World Athletics had reduced the number of 
false starts by two thirds. One study that looked at different false start
rules that IAAF has imposed from 1997 to 2009 and found that the reaction 
times when there was less leniency improved by a statistically significant amount
\citep{haugen2013effect}. This was desirable for World Athletics as false starts can 
make already long track meets more tedious, both for athletes and viewers watching the 
television broadcast.  Since 2011, IAAF has instituted the one false start rule
\citep{iaaf2009falsestart}.  However, we chose not to include women's data in this
paper due to the numerous papers citing difference in reaction times in men and
women's reaction times: \citep{lipps2011implications, babicc2009reaction,
panoutsakopoulos2020gender}
%Fix quotations
%Dont need the titles will add captions, use width =5, height = 3 in aspect ratio

\begin{figure}[tbp]
  \centering
  \includegraphics{Finals_Pooled_Boxplot}
  \label{fig:Boxplots}
  \caption{The reaction times from 2019 to 2022 for the men's final (top) and
  the men's semifinals and finals (bottom).}
\end{figure}

@@

\jy{set up the models/methods using the notations from last section; cite the
  right references.}

\subsection{GLMM}
%Include latex math symbols for the equations
The R package lme3 allows various linear mixed effects models to be created, and 
was the primary modeling package used for our data analysis \citep{lme4}.
A gamma mixed linear effects model was chosen as the best model to model the data.
Many models were tried and compared to determine what the strongest model.  For
a fixed data set such as the men's pooled data including 2022, the models with
the venue effect included performed better than the models without the venue
effect for each set of comparable models; as indicated by the lower AIC and
higher log likelihood.  Thus, the venue effect is significant and should be
included in the model.  It can also be established that a gamma distribution
better models the data than linear data from comparing the linear model with
venue effect and them gamma model with a venue effect.  The final decision that
was made in model selection was to include link = log in the model.  Despite
doing very little to alter the AIC or log likelihood, the log scale was included
in order to try and better deal with high outliers. %Include Residuals plot?

Once the Gamma mixed effects model was chosen, it needed to be decided which combination
of effects to use.  It was possible to include a venue effect, a batch effect, 
or a year and batch effect.  The batch effect is very
difficult to explain or interpret as it means that some heats of races were either
faster or slower and that everyone in the batch reacted similarly.  Although
it is possible that one athlete reacting faster could lead to others reacting
similarly, this does not make much practical sense as the difference between
heats that had fast and slow reaction times are all within 0.12 seconds of each
other.  It is much easier however to explain the venue effect.  The technology
that World Athletics has used has differed since 1999 and theoretically the
reaction time data from 2022 should be the best as the timing system should be
state of the art.  Any other weather or climate related factors such as humidity,
precipitation, elevation, etc. would all be negated when looking at the batch
effect, but are all factors that could potentially impact the venue effect.
Thus, we decided to inclue only the venue effect in the GLMM, which is presented
as:
\begin{equation}
  \label{eq:GLMM}
  Y_{ij} = \beta_0 + \gamma_i + \epsilon_{ij},
\end{equation}
where $Y_{ij}$ is the response variable for observation $j$ in year $i$, 
$\beta_0$ is the intercept, $\gamma_i$ is the random effect term for year $i$, 
and $\epsilon_{ij}$ is the error term for observation $j$ in year $i$.

%yij denotes ith years and j is observation in year
%yij will have clustered i and we use random effect to account for this
%When model is presented talk about how I modeled yij
%Zij where player is i and cluster, group is year and j
%Test hypothesis that reaction times are the same vs that they are not

>>>>

\jy{The bib file needs quality control. Practice my writing tips in the
  stat-wriging notes: chapter 2 on latex/bibtex.}

\jy{Go over the reference section to see what need to be changed in the bib source}

@@

After coming in third in the 110 meter hurdles at the United States Track and 
Field Championships in Eugene Oregon at the end of June, Devin Allen was expected to
be a contender for the World Championships, which were scheduled for two months 
later, also in Eugene. At the World Championships, Allen made it through the 
heats and semifinals to reach the finals while posting reaction times of 0.123
in the heats and 0.101 in the semifinal.  Allen, who attended the University
of Oregon (Location of US World Championship and the World Championship) was
considered to be a favorite for the 110 meter hurdle after running 12.84 seconds
in early June; only 0.04 seconds away from the world record \citep{wa2022preview}.
Thus his disqualification by reacting in 0.099 seconds instead of 0.01 was met
with shock from both Allen and fans alike.  NBC, who was broadcasting the
championships in the United States, later uploaded a video to their YouTube
channel that show the disqualification and can be found here:
\url{https://www.youtube.com/watch?v=D6NXTMo-1yM}. The crowd erupted with jeers when
they found out that Allen was disqualified, angry at the result.  Had Allen 
reacted just 0.001 seconds slower he would have been allowed to compete and been 
spared waiting another two years until the next World Track and Field Championship.
An investigation into reaction times is also important because reaction times have
been shown to affect the overall time of the race \citep{delalija2008reaction}.
Thus, faster reaction times lead to faster overall times; which is the goal of
every sprinter who steps onto the track.

Allen's disqualification has been discussed extensively in online communities
such as \url{www.LetsRun.com}, a website that is part message board and part
news. The message board,
which functions similarly to Reddit is very active during important running
events, which include during the finals of major events such as
the Olympics and the World Track and Field Championships.  However, it is the
\jy{cite these blog articles formally}
news section of the website that generated articles such as ``Was Devon Allen
Screwed? There's At Least a 99\% That He Was" and ``The Data Keeps Pouring In and
It Continues to Look Bad For World Athletics and Great For Devon Allen".  Robert
Johnson, who created and runs the Letsrun website argued that there was an error
with the timing equipment at the 2022 Track and Field World Championship and
cited graphs, statistics such as median reaction time, and compared the reaction
times at the 2022 Track and Field World Championship \citep{johnson2022devon}.

Reaction time in sprint and hurdle events has been studied by many authors.
\citet{pain2007sprint} looked at
reaction times for nine male athletes and analyzed their reaction times.  The
fastest athlete of the nine, had on average (even with their two fastest time 
removed) a reaction time of 0.087 seconds and a standard deviation of 0.004
seconds.  Two other athletes were able to react similarly when under certain
muscular tightening conditions.  In a study commissioned by IAAF,
\jy{define acronym at first occurrence}
\citep{komi2009iaaf} recommended that the disqualification 
barrier be decreased to allow sprinters to
react faster, and suggested 0.08 or 0.085 seconds as possible new thresholds.
Additionally, it was recommended to IAAF that high speed cameras be used to
change the reaction time criteria from pushing off on the block to body
movement. This could provide a more accurate representation of 
reaction time, and thus the possibility of a false disqualification would be lowered.
The possibility of using a video-based medium has been proved to be a reliable
measure of reaction time \citep{mudric2015evaluation}.

\jy{Could be condensed}
\citet{pilianidis2012start} found
significant differences between reaction times at several world championships
between 1997 and 2009 for the men's 110 meter hurdles.  Another important
conclusion was that for reaction times did not decrease over the twelve years
covered by the research in the study \citep{pilianidis2012start}.  This contradicts
what was found in a 2021 study that concluded the year of competition has a
significant effect on reaction times \citep{zhang2021correlation}.  Their
analysis looked at data from the 2011-2019 World Championships across multiple
sprint events and concluded that from 2011 to 2019, reaction times got
faster. However, not  all studies have been in favor of a decrease in the
reaction time barrier. \citet*{brosnan2017effects} in 2016 argue for increasing 
the reaction time barrier to 0.115 seconds, and cited evidence from European 
Championships and World Championships in their article.  A study looking 
at the 2008 Olympics likewise supported the conclusion that athletes could not
react in as little as 0.1 seconds \citep{lipps2011implications}.  Another study
concluded that visual reaction times to the starting gun may be faster by 0.007
seconds than the IAAF method, which is from force on block starts 
\citep{holmes2018method}.  While this may seem trivial, Allen was disqualified
only by 0.01 seconds. It is also worth noting that one study looking at data 
from various IAAF meets found that the fastest reaction times for men were found
between the ages of 26 and 29 \citep{tonnessen2013reaction}.  This is notable 
because Allen was 27 years old at the 2022 Track and Field World Championship 
and in his physical peak.

\jy{language needs to be formalized and condensed.}
The objective to this paper is to analyze the reaction times at the 2022
World Track and Field Championship to check for any abnormalities, and we approach
this two different ways.  Our first method is to try to model the data, which is
the reaction times of every non-disqualification since the 1999 World Championships. 
The model we use is a gamma linear mixed-effects model with a random effect
accounting for the venue (or year) effect. 
The second method is to compare data for athletes who have competed
at additional competitions to see how their reaction times compare from previous
competitions to this one. First we can look at the reaction times of athletes who
competed at the World Track and Field Championship in both 2019 and 2022 and
compare the times, and secondly we can look at how athletes performed compared
to how they did at their national/qualifiers meet. In order to reduce any chances
of athletes improving reaction time, we can also look at data from qualifying meets.
Almost every athlete had to qualify by proving they are elite at the national level, 
usually by coming in at least fourth. Thus, data was compiled on athletes who competed
at the 2022 World Track and Field Championships on what these athletes also ran 
at their country's national championship. If there was any argument that athletes
have improved their reaction time from 2019 to 2022, it should now be considered
nearly null, as the national competitions took place between June and July 2022 
and the World Championships were in August. The equipment used to measure reaction
time cannot be guaranteed to be the same from national to international
competition, but the time span between meets has been reduced considerably.

The rest of this paper is as follows: Section~\ref{sec:Data} describes how data was
collected and begins to detail the generalized linear mixed-effect (GLMM)
model which is further developed in the
Section~\ref{sec:Methods}.  The signed-rank test analysis for the same athletes is
also included in Section~\ref{sec:Methods}.  Section~\ref{sec:Results}
section provides the probabilities from the methods section and the paper ends 
by questioning the reaction time barrier.  Lastly, section~\ref{sec:Discussion}
discusses the impact and limitations of the paper.

@@

There are two types of data that is explored throughout this paper: data from
1999-2022 for every world championship, and data from athletes who competed
at both the 2019 and 2022 World Championships and 2022 national competitions
and the 2022 World Championship.  The first data set is used for the GLMM portion
of the paper, while the second data set is used for the rank based comparisons.

\subsection{World Championship}

\begin{figure}[tbp]
  \centering
  \includegraphics{Finals_Pooled_Boxplot}
  \caption{The reaction times from 2019 to 2022 for the men's final (top) and
  the men's semifinals and finals (bottom).}
  \label{fig:Boxplots}
\end{figure}

Data was taken from the World Athletics and covers the men's 110 meter hurdles 
and the women's 100 meter hurdles from 1999 to 2022.  The variable of prediction 
interest is reaction time, and the predicting variables are year, stage of 
competition (heats, semifinal, final), total time, and gender.  
The data can be found in the appendix~\ref{sec:Appendix}. Factors such as name, 
country, and identification number were not considered important for the 
purposes of this research besides the times for Devon Allen and the other 
United States athletes mentioned above.  The variable ``Stage"
refers to whether the observation occurred during the ``heats" (Preliminary round) which
is denoted by ``H" in the data, ``S" for ``semifinals" (Of which there are either 2 or 3 
heats each year), and ``F" for finals. The variable ``Gender" has levels ``M" and ``F" 
which stand for ``male" and ``female" respectively.  The variable ``Batch" takes a 
numerical value and every heat of every round from 1999 to 2022 was assigned it's
own number.  This was added because of suspected issues that reaction time may be
correlated with other runners reacting quickly (one runner reacting quickly may
result in other runners reacting quickly).  There is no statistical significance to
how the batches were labeled, they were ordered chronologically by meet but
reverse chronologically by year (Batch 1 is a preliminary Heat of 2022 and Batch
111 corresponds to the 1999 finals). Included are
box plots of the reaction times of the men's finals and the men's semifinals and
finals pooled together.  We will refer to the latter as the ``pooled data"
throughout the paper.  The heats or preliminary rounds were not studied because
athletes react faster as the competition progresses.  Athletes react faster in
the finals than the semifinals which are faster than the heats 
\citep{zhang2021correlation}.  If we want to determine if Allen's reaction time
is reasonable, then we thus need to look at the men's final data, with the men's
pooled data being included in order to increase the sample size.  In 2022, there
were only five data points due to two disqualifications and one athlete who 
qualified but did not compete.
The data that looked at in the paper is best visualized using a box plot as show 
in Figure~\ref{fig:Boxplots}.  It is clear that the 2022 box plot for both the
men's finals and semifinals data is lower (indicating a faster reaction time)
than in previous years.  The median for the men's pooled data in 2022 was 0.129
and the median for the men's final data in 2022 was 0.126.  For comparison,
Brosnan found reaction times of 0.156 seconds for men when looking at data from
1999 to 2014 \citep{brosnan2017effects}.

Something that is extremely important to note in the data is the inconsistency 
of World Athletics in terms of how they penalize false starts.  From 2007--2009,
the IAAF (the former name for World Athletics, and the governing body for the 
World Track and Field Champions) instituted a rule change that allowed one 
warning false start before a sprinter was disqualified \citep{iaaf2009falsestart}  This 
essentially gave sprinters a second chance and reduced the penalty for a false 
start.  There were 18 male and 7 female false starts at the 2007 World 
Championships, 18 male and 7 female starts at the 2009 World Championships. 
Starting in 2011, the rule was scrapped, and the old policy which automatically 
disqualifies runners who false start was reinstituted.  The effect? At the 2011 
World Championship there were 6 male disqualifications and 4 female 
disqualifications \citep{iaaf2009falsestart}. By returning to the harsher policy and 
cracking down on false starts, World Athletics had reduced the number of 
false starts by two thirds. One study that looked at different false start
rules that IAAF has imposed from 1997 to 2009 and found that the reaction 
times when there was less leniency improved by a statistically significant amount
\citep{haugen2013effect}. This was desirable for World Athletics as false starts can 
make already long track meets more tedious, both for athletes and viewers watching the 
television broadcast.  Since 2011, IAAF has instituted the one false start rule
\citep{iaaf2009falsestart}.  However, we chose not to include women's data in this
paper due to the numerous papers citing difference in reaction times in men and
women's reaction times: \citep{lipps2011implications, babicc2009reaction,
panoutsakopoulos2020gender}

@@

The World Championship data and the data beyond the World Champisonship were
analysed with GLMM and rank-based approaches, respectively.

\subsection{GLMM}
Based on an exploratory analysis, the GLMM model that we chose was a gamma
mixed-effect model for the reaction times from the World Championship. The
normal error model on the log-scale was found to be inferior to the gamma model.
Let $Y_{ij}$ be the reaction time of observation~$j$ in year~$i$.
Conditional on a venue effect $z_i$ of year~i,  $Y_{ij}$ is modeled by a
gamma distribution with mean
\[
\mu_{i} = \alpha + z_i,
\]
and variance
\[
\sigma_i
\]
...
\jy{need to add assumptions ...}

The R package \texttt{lme4} allows various linear mixed effects models to be created, and 
was the primary modeling package used for our data analysis \citep{lme4}.
A gamma mixed linear effects model was chosen as the best model to model the data.
Many models were tried and compared to determine what the strongest model.  For
a fixed data set such as the men's pooled data including 2022, the models with
the venue effect included performed better than the models without the venue
effect for each set of comparable models; as indicated by the lower AIC and
higher log likelihood.  Thus, the venue effect is significant and should be
included in the model.  It can also be established that a gamma distribution
better models the data than linear data from comparing the linear model with
venue effect and them gamma model with a venue effect.  The final decision that
was made in model selection was to include link = log in the model.  Despite
doing very little to alter the AIC or log likelihood, the log scale was included
in order to try and better deal with high outliers. %Include Residuals plot?

Once the Gamma mixed effects model was chosen, it needed to be decided which combination
of effects to use.  It was possible to include a venue effect, a batch effect, 
or a year and batch effect.  The batch effect is very
difficult to explain or interpret as it means that some heats of races were either
faster or slower and that everyone in the batch reacted similarly.  Although
it is possible that one athlete reacting faster could lead to others reacting
similarly, this does not make much practical sense as the difference between
heats that had fast and slow reaction times are all within 0.12 seconds of each
other.  It is much easier however to explain the venue effect.  The technology
that World Athletics has used has differed since 1999 and theoretically the
reaction time data from 2022 should be the best as the timing system should be
state of the art.  Any other weather or climate related factors such as humidity,
precipitation, elevation, etc. would all be negated when looking at the batch
effect, but are all factors that could potentially impact the venue effect.
